---
title: "Data Science Lab in R - Lab Two"
author: "Ishanvi Verma"
date: "2025-06-22"
output: pdf_document
---

# Short Answer Questions

**For the following questions, please response in 3-4 sentences.**

## Question 1

**Many people expect a coin flip to be 50-50 (i.e. a binomial distribution). Read this article about a (fairly extreme) [coin toss experiment](https://www.scientificamerican.com/article/scientists-destroy-illusion-that-coin-toss-flips-are-50-50/). Describe your thoughts.**

I think it was interesting how through 350,707 tosses they were able to realize about a 1% bias and rather then thinking that with more tosses it would have approached the exact 50%. It is interesting how they investigated this as a bias and connected the concept to physics of how there was a greater chance of it landing on the same side than actually flipping. One thing I would ask is if they truly got rid of the bias in algorithm since not everyone could have truly done a coin flip maybe some people just threw the coin up... anyways it is possible after 350,707 tosses for people to have gotten lazy.

## Question 2

**In your own words, describe how to take a bootstrap sample.**

You have take a sample of some observations maybe in this case heights of a girl friend group to find the average height between a group of girls. Then you can randomly select each height of the girl from a friend group. After choosing every individual data point from the original sample replace it. Repeat the following process n times. Record the mean of the n resampled vales. Then keep going that for like x number of times and use those the x results to calculate the final mean by averaging the other mean values from the intervals of resampling.

## Question 3

**When do you think we should rely on parametric assumptions (e.g. using a known distribution to approximate something) and when should we rely on simulation-based practices (e.g. something like bootstrap)?**

We should use parametric assumptions when we are sure that the data follows a normal distribution or when the sample size is large enough for CLT to apply (so it is more prone to accurately depict a population). We can use simulation-based practice like bootstrapping to compensate for a small sample size and when we don't want to make as many assumptions. Or we can also use simulation-based practices when trying to find medians since it is harder with larger data sets.

## Question 4

**Describe the importance of EDA before analyzing the data.**

EDA helps with summarizing data and catching any errors before running any formal analysis. Without it one can risk making misleading conclusions off incorrect data. EDA can also help analyze trends which is why it is important because it helps with getting the most out of the data and drawing conclusive statements. EDA also helps with analyzing trends even when there are outliers.

## Question 5

People also assume that the gender ratio (the ratio of men to women) is 1:1. Skim this [article](https://ourworldindata.org/gender-ratio) and describe your thoughts.

This really shows a good example of how they are proving correlation is a causation by taking many different factors into aspect of why the gender ratio is not 1:1. Another idea I wanted to note which I do not think was noted, is that in many third-world nations finding out the gender is illegal because at a time many families would have abortion when they found out their baby was female which could also cause the imbalance to the 1:1 ratio. These scientists looking further out from women mortality rates at 10-15 weeks both as a baby and mother, as well as the rate at women being born shows how one conclusion is not always from one investigation but many combined – trends from more than one data distribution.

# Coding Questions

**For each question below, if you create a variable or an R object, please print out the variable after creating.**

**Click "Knit" above regularly to generate a pdf of this document. If you only knit the document at the end, it's much more difficult to troubleshoot.**

## Section 1: Loading packages

**Load the `tidyverse` library.**

```{r, message=FALSE}
library(tidyverse)

```

**Load the `gapminder` library.**

```{r, message=FALSE}
library(gapminder)

```

**We will be working with the gapminder dataset, so use `data(...)` to load it into your environment.**

```{r}
data("gapminder")
```

**Use the `?`** **operator, and write down what are the columns in the dataset.**

The columns are "country", "continent", "year", "gdpPercap", "lifeExp", and "pop.

**Use the `head()` function to print out the first 5 rows.**

```{r}
gapminder |>
  head(n=5)
```

**Use the `tail()` function to print out the last 3 rows.**

```{r}
gapminder |>
  tail(n=5)
```

## Section 2: `dplyr` pipelines

**TASK 1: Find the average life expectancy and its standard deviation across continents in the year 1997.**

**First, use `filter()` to create a new dataframe called `gapminder97` with just 1997 data.**

```{r}
gapminder97 <- gapminder |>
  filter(year==1997)
gapminder97  
```

**Then, use `summarize()` to find average life expectancy and its standard deviation for ALL continents.**

```{r}
gapminder97 |>
    summarize(mean_lifeExp = mean(lifeExp),
              sd_lifeExp = sd(lifeExp)) 


```

**Add `group_by()` to get your statistics separated by continent. (Be conscious of where this goes in the pipeline!)**

```{r}
gapminder |>
  group_by(continent) |>
    summarize(mean_lifeExp = mean(lifeExp),
              sd_lifeExp = sd(lifeExp))
```

**TASK 2: Try to change the code above to get the median population separated by continent AND year.**

```{r}
gapminder |>
    group_by(continent, year) |>
      summarize(median_population = median(pop)) 

```

**TASK 3: Find the Asian continent with the highest GDP per capita in 2002.**

**First, `filter()` to get only data on Asian countries from 2002. Call this dataframe `Asia2002`.**

```{r}
Asia2002 <- gapminder |>
  group_by(continent, year) |>
  filter(continent == "Asia", year == 2002)
```

**Then, `arrange()` the data in descending order. (Note that `arrange()` defaults to ascending).**

```{r}
Asia2002 |>
  arrange(desc(gdpPercap))

```

**Use `select()` to only show the country, continent, year and gdp per capita columns.**

```{r}
Asia2002 |>
  arrange(desc(gdpPercap)) |>
    select(country, continent, year, gdpPercap)
```

**Use `head()` to print only the first row.**

```{r}
Asia2002 |>
  arrange(desc(gdpPercap)) |>
    select(country, continent, year, gdpPercap) |>
      head(n=1)
```

**TASK 4: Find the country with the highest real GDP in 1972 in Europe.**

**Note that** $GDP Per Capita = \frac{Real GDP}{Population}$**. Use `mutate()` to create a realGDP column.**

```{r}
Europe72 <- gapminder |>
  filter(year == 1972, 
         continent == "Europe") |>
    mutate(realGDP = gdpPercap * pop)
```

**Repeat similar steps from Task 3.**

```{r}
Europe72 |>
  arrange(desc(realGDP)) |>
  select(country, continent, year, gdpPercap, realGDP)|>
  head(n=1)
```

## Section 3 : `ggplot()`

**TASK 1: Create a histogram for life expectancy in 1952, with color representing the continent.**

```{r}
gapminder |>
  filter(year == 1952) |>
  ggplot(aes(x=lifeExp, fill = continent))+
  geom_histogram()
```

**TASK 2: Create a histogram for life expectancy in 1952, faceted by continent using `facet_wrap()`. Which plot do you prefer: Task 1 or Task 2? Why?**

```{r}
gapminder |>
  filter(year == 1952) |>
    ggplot(aes(x=lifeExp, fill = continent))+
    geom_histogram()+
    facet_wrap(vars(continent))
```

**TASK 3: Create a scatterplot of year vs. population for a country of your choosing.**

```{r}
gapminder |>
  ggplot(aes(x=year, y=pop))+
  geom_point()
```

**TASK 4: Create a bar chart of year vs. population for the same country using `geom_col()`. Which plot do you prefer: Task 3 or Task 4? Why?**

```{r}
gapminder |>
  ggplot(aes(x=year, y=pop))+
  geom_col()
```

Written answer here.

**TASK 5: Using your plot from Task 4, add a `geom_line()`.**

```{r}
gapminder |>
  ggplot(aes(x=year, y=pop))+
  geom_col()
  geom_line()
```

**TASK 6: Recreate the Hans Rosling plot for the year 1992 (no animations!).**

```{r}
gapminder |>
  filter(year == 1992) |> 
  ggplot(aes(x=gdpPercap,
             y=lifeExp,
             size = pop,
             color = continent)) + 
  geom_point()
```

## Section 4: PDFs and CDF

**Recall the game from your first quiz: Consider the following game. Roll a fair 8-sided die. If you roll an even number, you lose \$5. If you roll 5 or 7, you win \$1. If you roll a 3, you get nothing. If you roll a 1, you win \$10. What's the expected value of this game? Show your work.**

**TASK 1: Create a dataframe called `pdf` that has two columns: one called `x` for the outcomes and one called `prob` for the respective probabilities. Use this dataframe and `dplyr` functions to calculate expected value.**

**First, create the dataframe:**

```{r}
pdf <- data.frame(x = c(-5,1,0,10),
                  prob = c(0.5, 0.25, 0.125, 0.125))
```

**Then, use `mutate()` and `summarize()` to calculate the expected value.**

```{r}
pdf|>
  mutate(outcome = x*prob) |>
    summarize(weighted_mean = sum(outcome))
```

**Use `weighted.mean()` to check your answer.**

```{r}
weighted.mean(pdf$x, pdf$prob)
```

Use `ggplot` to visualize the probability distribution.

```{r}
pdf|>
  mutate(outcome = x*prob) |>
    ggplot(aes(x=x, y=prob, fill=outcome))+
    geom_col()
```

**TASK 2: Find the cdf of this distribution.**

**Hint: use** `mutate()` **and the function `cumsum()`** .

```{r}
cdf <- pdf |>
  mutate(prob = cumsum(prob))
cdf
```

## Section 5: Probability Simulations

**TASK 1: Women's height tend to be normally distributed. Let's plot it!**

**First, find the average and standard deviation of American women (in inches) and save it as two variables.**

```{r}
women_ht_avg <- 63.6
women_ht_sd <- 2.5
```

Then, create a dataframe of x-values (from 50 to 75, at 0.1 intervals).

```{r, echo=TRUE, results='hide'}
x = (seq(from=50, to=75, by=0.1))
df = data.frame(x)
df
```

Use `mutate()` to calculate the probabilities of those x-values.

```{r}
df <- data.frame(x = x) |>
  mutate(probability = dnorm(x, mean = women_ht_avg, sd = women_ht_sd))
df
```

Plot these points (using a scatterplot).

```{r}
df |>
  ggplot(aes(x=x, y=probability))+
  geom_point()
```

Add a vertical line at your own height on the plot.

```{r}
df |>
  ggplot(aes(x=x, y=probability))+
  geom_point()+
  geom_vline(xintercept = 60, color = "lightpink", linewidth =1)

```

**TASK 2:** **Create a plot of the pdf for 10 rolls of a die with success defined as rolling a prime number.**

```{r}
x <- 0:10
data.frame(x=x) |>
  mutate(prob=dbinom(x, size=10, prob=0.5)) |>
    ggplot(aes(x=x, y=prob))+
    geom_col()
```

**TASK 3: Are you more likely to get 5 heads in 10 tosses of a fair coin or 50 heads in 100 tosses of a fair coin?**

```{r}
toss1 = dbinom(5, size=10, prob=0.5)
toss2 = dbinom(50, size=100, prob=0.5)
toss1
toss2
```

**TASK 4: What's the probability of getting at most 50 heads in 100 tosses of a fair coin?**

```{r}
pbinom(q=50, size=100, prob=0.5)
sum(dbinom(x=0:50, size=100, prob=0.5))
```

**TASK 5: Create a plot of the pdf for a Poisson distribution with lambda = 1. Use** `geom_point` **AND** `geom_line` **.**

**Hint: use integers from 0 to 10 as your x-values**

Written answer here

```{r}
lambda_val <- 1
x_vals_plot <- 0:10

pdf_data <- data.frame(
  x = x_vals_plot,
  probability = dpois(x_vals_plot, lambda = lambda_val)
)

ggplot(pdf_data, aes(x = x, y = probability)) +
  geom_point() + 
  geom_line()  
```

## Section 6: Comparing Distributions

**TASK 1: Assuming a large number of trails (n = 100+), for what values of p does the binomial distribution look very similar to a normal distribution? Use plots of the pdf to demonstrate.**

Written answer here.

```{r}
x <- 0:1000
data.frame(x=x)|>
  mutate(prob = dbinom(x,size=1000,prob=0.5)) |>
  ggplot(aes(x=x, y=prob))+
  geom_point()
```

**TASK 2: Assuming a large number of trails (n = 100+), for what values of p does the binomial distribution look very similar to the Poisson distribution above? Use plots of the pdf to demonstrate.**

Written answer here.

```{r}
data.frame(
var <- c(0:1000),
  binom <- dbinom(var,1000,0.05)) |>
  ggplot(aes(x=var,
             y=binom))+
  geom_point()

```

**TASK 3: For what values of lambda does a Poisson distribution look very similar to a normal distribution. Use plots of the pdf to demonstrate.**

Written answer here.

```{r}
x = 0:1000
probab <- dpois(x,
                lambda = 500)
data_poisson <- data.frame(x=x,
                           y=probab)
data_poisson |>
ggplot(aes(x=x,
           y= probab))+
  geom_point() + geom_line()
```
